---
title: "Model diagnostics guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model diagnostics guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

RTMR ships a full set of diagnostics around the RTMDx-replicating pipeline. This vignette is the roadmap: which tool answers which question, and — since each stage's answer changes how the next stage should be run — the order in which to use them.

The order, and the reasoning behind it:

1.  **Design decisions** (before fitting): exclude impossible areas, fix the seed, stabilize the culling. These cannot be retrofitted.
2.  **Does the model describe its own data?** (`check_fit()`) — coefficients from a model that cannot reproduce its own data are not worth interpreting, so absolute fit comes first.
3.  **Can the cells be treated as independent?** (`moran_rtm()`, `correlogram_rtm()`) — the answer decides *how everything downstream is reported*: which standard errors to use, and whether resampling must be spatially blocked.
4.  **Is the selection trustworthy?** (`robust_rtm()`, `stability_rtm()` or `split_events()` + `refit_rtm()`) — honest uncertainty for the RRVs, using the dependence verdict from step 3.
5.  **Does it predict new events?** (`validate_rtm()`, temporal `refit_rtm()`) — the payoff question, asked last because a model that fails steps 2–4 predicts well only by accident.
6.  **Sensitivity analyses** — tuning-choice robustness, reported alongside the primary results.

Theory and interpretation live in the two specialized vignettes — `vignette("spatial-inference")` for step 3 and the dependence side of step 4, `vignette("post-selection")` for the selection side — this guide shows the workflow end to end.

```{r setup}
library(RTMR)
library(sf)
```

## A worked example

The simulated setting of `vignette("RTMR")`: crimes cluster around 15 bars, 10 parks are irrelevant.

```{r simulate}
set.seed(42)
size <- 3000

boundary <- st_sf(geometry = st_sfc(
  st_polygon(list(rbind(
    c(0, 0), c(size, 0), c(size, size), c(0, size), c(0, 0)
  ))),
  crs = 32610
))

rand_pts <- function(n) cbind(runif(n, 0, size), runif(n, 0, size))
as_sfc <- function(m) st_sfc(lapply(seq_len(nrow(m)), \(i) st_point(m[i, ])), crs = 32610)

bars <- rand_pts(15)
parks <- rand_pts(10)

gen_crimes <- function() {
  ev <- do.call(rbind, lapply(seq_len(nrow(bars)), function(i) {
    n <- rpois(1, 8)
    cbind(rnorm(n, bars[i, 1], 100), rnorm(n, bars[i, 2], 100))
  }))
  ev <- rbind(ev, rand_pts(30))
  as_sfc(ev[ev[, 1] >= 0 & ev[, 1] <= size & ev[, 2] >= 0 & ev[, 2] <= size, ])
}
crimes <- gen_crimes()
crimes2 <- gen_crimes() # a second period, held out for step 5

bars <- as_sfc(bars)
parks <- as_sfc(parks)
```

## Step 1: design decisions, before fitting

Three choices belong *before* the fit, because no later diagnostic can undo them:

-   **Mask impossible areas.** Cells where events cannot occur (water bodies, restricted land) are structural zeros that dilute the model; pass them as `exclude` so the analysis is restricted to places where events are possible. (None here — the toy area is all "land".)
-   **Fix the seed.** The culling step randomizes its cross-validation folds, as in RTMDx; `set.seed()` makes the whole run exactly reproducible.
-   **Stabilize the culling.** `cull_repeats = 5` averages the culling cross-validation over five independent fold assignments, so which variables survive does not hinge on one lucky draw.

```{r fit}
set.seed(1)
fit <- rtm(
  outcome = crimes,
  factors = list(bars = bars, parks = parks),
  boundary = boundary,
  cell_size = 150,
  block_length = 300,
  operation = "proximity",
  cull_repeats = 5,
  verbose = FALSE
)
fit
```

Remember that the p-values in this table are a selection sanity check, not inference — honest uncertainty comes in step 4.

## Step 2: does the model describe its own data?

`check_fit()` answers the absolute-fit question with pseudo-R² values and posterior-predictive checks, and its plot draws the two standard count-model visuals:

```{r check-fit}
set.seed(2)
cf <- check_fit(fit)
cf
```

```{r check-fit-plot, fig.width = 8, fig.height = 4}
plot(cf)
```

What to look for, in order of importance:

-   **`zero_share` flagged** → the mask from step 1 is incomplete (more structural zeros to exclude) or the process is genuinely zero-inflated. Fix and refit before proceeding.
-   **`dispersion` flagged under a Poisson fit** → residual overdispersion; the BIC contest usually hands this to the negative binomial automatically, so a flag here deserves investigation.
-   **Rootogram bars off the zero line / Q-Q off the diagonal** → the count range where the model fails, localized.
-   **Pseudo-R²** calibrates *how much* is explained — expect small values on sparse cell counts; compare models on the same data, not against OLS conventions.

Here all three checks pass and the visuals sit clean, so the model is worth interrogating further.

## Step 3: can the cells be treated as independent?

Everything the pipeline reports — p-values, the BIC penalty, the culling cross-validation — assumes independent cells. `moran_rtm()` tests that assumption on the residuals:

```{r moran}
set.seed(3)
moran_rtm(fit)
```

No detectable residual clustering. Had this flagged (it does when a spatially structured factor is missing — see the worked failure in `vignette("spatial-inference")`), the follow-up chain is:

1.  `correlogram_rtm(fit)` — estimate how far the residual correlation reaches;
2.  use its suggested `cluster_size` in step 4's `robust_rtm()`, and switch `stability_rtm()` to `resample = "block"`;
3.  optionally refit the selected specification with a spatial smooth (`mgcv::gam` with `s(x, y)`) as a *sensitivity* check on the RRVs themselves.

The correlogram is cheap, so it is worth running even on a pass — "no range detected" is itself a reportable result:

```{r correlogram}
set.seed(4)
cg <- correlogram_rtm(fit, nsim = 199)
cg$range
```

## Step 4: is the selection trustworthy?

Two distinct concerns, two tools, both reported against the *unchanged* point estimates:

**Dependence-honest intervals.** `robust_rtm()` replaces the model-based standard errors with spatially clustered ones. When step 3 passed, robust and model-based p-values agree and reporting the robust ones costs nothing; when step 3 flagged, they diverge and the robust ones are the defensible choice. Use the correlogram's range (or the default) for `cluster_size`:

```{r robust}
rob <- suppressWarnings(robust_rtm(fit, cluster_size = 600))
rob$selected[, c("factor", "rrv", "ci_lower", "ci_upper", "p_robust", "p_model")]
```

(The few-clusters warning is the 3 km toy area's fault — a city-scale grid gives ample clusters.)

**Selection-aware estimates.** The RRVs were *selected*, so they are biased away from 1 (the winner's curse) and their p-values are conditional on the selection. With plentiful events, split and refit (`split_events()` + `refit_rtm()`) gives exact selection-free inference; with scarce events, bootstrap the pipeline instead — full sample, selection frequencies, bagged RRVs:

```{r stability}
set.seed(5)
stability_rtm(fit, nboot = 50, verbose = FALSE)
```

Bars is selected in essentially every resample with a bagged RRV close to the reported one (a stable finding, mildly inflated at most); parks' low frequency is what exposes a would-be spurious factor. The decision rule between the two tools — and how they compose with step 3 — is in `vignette("post-selection")`.

## Step 5: does it predict new events?

The decisive test is data the model has never seen — typically the next time period:

```{r validate}
val <- validate_rtm(fit, crimes2)
val
```

PAI says how concentrated the captured events are relative to the flagged area; AUC summarizes cell-level discrimination; observed/expected checks calibration when the periods have equal length (for rate models, pass `new_offset` for denominator-aware capture and a duration-free calibration check). And re-estimating the frozen specification on the new period asks whether the *effects* transport:

```{r temporal-refit}
refit_rtm(fit, crimes2)
```

A temporal change here mixes selection bias with genuine drift — the thinned split of `vignette("post-selection")` separates the two.

## Step 6: sensitivity analyses

Last, show the conclusions do not hinge on tuning choices. The usual suspects:

-   **Density threshold**: refit with `density_quantile = 0.95` versus the RTMDx mean + 2 SD rule when density operationalizations were used.
-   **Cluster size**: report `robust_rtm()` at a second, coarser `cluster_size` and show the intervals barely move.
-   **Smoothed offset** (rate models): `smooth_offset()` at two or three bandwidths, reporting RRV stability (see `vignette("RTMR")` for why raw counts stay the primary analysis).
-   **Spatial trend** (if step 3 flagged): the `mgcv` refit from `vignette("spatial-inference")`.

## When the fit is poor and you don't know why

The steps above assume each diagnostic points at its own remedy. The harder situation is a poor fit with *no obvious candidate* for what is missing. Two things help: the failure signature classifies the problem, and the residuals localize it.

**Read the signature first.** Which check failed narrows what kind of thing is missing:

| Signature | Likely cause | Next move |
|---|---|---|
| `zero_share` flagged | Structural: unmasked impossible areas, denominator coverage, geocoding | Fix `exclude`/data, refit |
| `dispersion` flagged, `moran_rtm()` quiet | Unobserved heterogeneity that is *not* spatially organized (often person-level; see below) | NB absorbs it: wider intervals, no bias — usually acceptable |
| `moran_rtm()` flagged | A missing *spatially structured* factor; `correlogram_rtm()`'s range estimates its footprint (300 m = micro-place feature, 2 km = neighbourhood process) | Hunt at that scale (below), refit with the new factor |
| Fits in-sample, fails `validate_rtm()` | Not a missing factor: instability or temporal drift | `stability_rtm()`, temporal `refit_rtm()` |

**Then let the residuals do the searching.** `plot(fit, what = "residual")` maps the Pearson residuals on a diverging scale: red cells host more events than the measured environment explains — the candidate locations of whatever is missing. Refitting our example *without* the bars layer shows how directly this works (crosses mark the withheld bars):

```{r residual-map, fig.width = 6, fig.height = 5.5}
set.seed(6)
fit_nobars <- rtm(
  outcome = crimes,
  factors = list(parks = parks),
  boundary = boundary,
  cell_size = 150,
  block_length = 300,
  operation = "proximity",
  verbose = FALSE
)
plot(fit_nobars, what = "residual", reset = FALSE)
plot(st_geometry(bars), add = TRUE, pch = 4, lwd = 2)
```

The workflow from here is deliberately qualitative: take the reddest locations (and the correlogram's scale estimate) to people with domain knowledge — what is *at* those places? — then encode the answer as a factor layer, refit, and rerun the chain from step 2. The `mgcv` spatial-smooth refit of `vignette("spatial-inference")` produces a smoothed version of the same map. This iterative use is how RTM is meant to work: the diagnostics turn "the model is bad" into "go find out what is at these coordinates, at roughly this spatial scale."

### When the missing factors are personal, not areal

Individual-level drivers (age, tolerance, using alone) reach an areal model only through *who is where*, which yields exactly two cases:

-   **Composition is not spatially organized.** Person-level variation then cannot bias any place-level coefficient; it surfaces purely as overdispersion, which the pipeline's negative binomial option absorbs (that is what the NB's mixing distribution *is*). Wider intervals, no bias — a limitation priced in, not a failure.
-   **Composition is spatially organized** (high-risk individuals concentrate in particular places — the realistic case). Then composition *is* an areal variable, whether measured or not: it triggers `moran_rtm()` like any missing spatial factor, and the remedy is to measure it at the cell level and adjust for it.

For the second case, cell-level *summaries of person-level data* (share of events involving people who used alone, mean age, % unstably housed) are usually better than coarse administrative proxies — finer resolution controls composition where it actually varies. Pass them as `covariates` with `role = "adjustment"`, with four cautions: (1) build them from an *external source or a prior period*, not from the same events being modeled, to avoid circularity; (2) stabilize sparse cells — proportions from one or two people are noise, so shrink towards the overall mean or aggregate coarser; (3) fine-grained aggregates of health data may face small-cell privacy rules; and (4) adjusting for composition changes the estimand — RRVs become *contextual* effects net of who is present, which is often exactly what is wanted, but say so. What no areal analysis can do is fully separate a place's causal effect from the people it attracts (the contextual-vs-compositional problem); that requires individual-level designs, and is a scope statement for the writeup rather than a fixable defect.

## Quick reference

| Order | Question | Tool | If it fails |
|---|---|---|---|
| 1 | Impossible areas? Reproducible? | `exclude`, `set.seed()`, `cull_repeats` | — (design; cannot be retrofitted) |
| 2 | Reproduces its own data? | `check_fit()` + `plot()` | zeros → extend `exclude` or suspect zero-inflation; dispersion → NB; then refit |
| 3 | Independent cells? | `moran_rtm()` → `correlogram_rtm()` | robust SEs in step 4, `resample = "block"`, mgcv sensitivity refit |
| 4 | Honest uncertainty? | `robust_rtm()`; `stability_rtm()` or `split_events()` + `refit_rtm()` | report bagged/refit estimates; drop low-frequency factors from findings |
| 5 | Predicts new events? | `validate_rtm()`, temporal `refit_rtm()` | revisit factors/spatial influences; check drift vs selection |
| 6 | Robust to tuning? | `density_quantile`, `cluster_size`, `smooth_offset()` | report the sensitivity range |

Two habits worth keeping throughout: `set.seed()` before every stochastic step (all of the diagnostics above use permutation, simulation, or resampling), and when a diagnostic fails, fix and *rerun the chain from that step down* — the stages are ordered precisely because each one's verdict shapes the next.

## References

-   Heffner, J. (2013). Statistics of the RTMDx Utility. In J. Caplan, L. Kennedy, & E. Piza, *Risk Terrain Modeling Diagnostics Utility User Manual (Version 1.0)*. Newark, NJ: Rutgers Center on Public Security.
-   Cameron, A. C., & Windmeijer, F. A. G. (1996). R-squared measures for count data regression models. *Journal of Business & Economic Statistics*, 14, 209–220.
-   Kleiber, C., & Zeileis, A. (2016). Visualizing count data regressions using rootograms. *The American Statistician*, 70, 296–303.
-   Dunn, P. K., & Smyth, G. K. (1996). Randomized quantile residuals. *Journal of Computational and Graphical Statistics*, 5, 236–244.
-   Chainey, S., Tompson, L., & Uhlig, S. (2008). The utility of hotspot mapping for predicting spatial patterns of crime. *Security Journal*, 21, 4–28.
