---
title: "Risk terrain modeling with RTMR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Risk terrain modeling with RTMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

Risk Terrain Modeling (RTM) diagnoses which features of the environment (bars, transit stops, vacant lots, ...) are associated with where events such as crimes concentrate, estimates *how far* each feature's influence reaches, and combines the selected factors into a map of relative risk. RTMR replicates the statistical procedure of the commercial RTMDx Utility (Heffner, 2013) and extends it with rate (offset) models and holdout validation.

The pipeline, in brief:

1.  A grid of cells is laid over the study area; the outcome events are counted per cell.
2.  Each risk factor is *operationalized* into candidate binary variables at a ladder of spatial influences: proximity ("within *d* of a feature") and/or density ("inside a high-concentration cluster at bandwidth *d*").
3.  A cross-validated, sign-constrained penalized Poisson regression culls the candidates, protecting against spurious correlations from testing many variables.
4.  Bidirectional stepwise regression by BIC selects the final factors and their optimal spatial influences, under both Poisson and negative binomial distributions; the best BIC wins. At most one variable — one operationalization at one distance — may represent each factor.
5.  The fitted surface, rescaled to its minimum, gives each cell's relative risk score, and each factor's exponentiated coefficient is its relative risk value (RRV).

```{r setup}
library(RTMR)
library(sf)
```

## Simulated example data

We simulate a 3 km × 3 km study area with two candidate risk factors — 15 bars and 10 parks — where crimes truly cluster around bars (within about two hundred metres) and parks are irrelevant. A working CRS in metres is assumed throughout; project your own data before modeling.

```{r simulate}
set.seed(42)
size <- 3000

boundary <- st_sf(geometry = st_sfc(
  st_polygon(list(rbind(
    c(0, 0), c(size, 0), c(size, size), c(0, size), c(0, 0)
  ))),
  crs = 32610
))

rand_pts <- function(n) cbind(runif(n, 0, size), runif(n, 0, size))
as_sfc <- function(m) st_sfc(lapply(seq_len(nrow(m)), \(i) st_point(m[i, ])), crs = 32610)

bars <- rand_pts(15)
parks <- rand_pts(10)

crimes <- do.call(rbind, lapply(seq_len(nrow(bars)), function(i) {
  n <- rpois(1, 8)
  cbind(rnorm(n, bars[i, 1], 100), rnorm(n, bars[i, 2], 100))
}))
crimes <- rbind(crimes, rand_pts(30)) # background noise
crimes <- crimes[crimes[, 1] >= 0 & crimes[, 1] <= size &
  crimes[, 2] >= 0 & crimes[, 2] <= size, ]

bars <- as_sfc(bars)
parks <- as_sfc(parks)
crimes <- as_sfc(crimes)
```

## Fitting a risk terrain model

`rtm()` runs the whole pipeline. Two spatial parameters matter most: `block_length`, the average block length of the study area (spatial influences are tested in block increments), and `cell_size`, the grid resolution (RTMDx recommends half a block).

```{r fit}
set.seed(1)
fit <- rtm(
  outcome = crimes,
  factors = list(bars = bars, parks = parks),
  boundary = boundary,
  cell_size = 150,
  block_length = 300,
  operation = "both", # test proximity and density variants
  increment = "half", # half-block distance ladder
  verbose = FALSE
)
fit
```

The model recovered bars as a risk factor and estimated its spatial influence, while (usually) discarding parks or assigning it a much weaker role. Per-factor settings are supported by passing a list instead of a bare layer, e.g. `factors = list(ops = list(data = ops, type = "protective", max_blocks = 2))`.

### Polygon factors and cell-level covariates

Factors need not be points. Line and polygon layers (parks, transit corridors) are supported for the *proximity* operationalization — distance is measured to the nearest edge, and cells inside a polygon are always exposed. Density is a kernel density of a point pattern, so requesting it for non-point geometry is an error; assign operations per factor when mixing geometry types:

```{r polygons, eval = FALSE}
factors = list(
  bars  = list(data = bars_pts, operation = "both"),
  parks = list(data = park_polygons, operation = "proximity")
)
```

### Masking places where events cannot occur

Most grid cells have zero events, and that is usually fine: Heffner (2013) shows the zero share expected under a Poisson process matches what typical data produce. But zeros of a different kind — *structural* zeros, cells where events are impossible (water bodies, an airfield, unpopulated industrial land) — do not belong in the model at all: they dilute the fit with places that carry no information about risk. Rather than absorbing them statistically (zero-inflated models), exclude them:

```{r exclude}
lake <- st_sfc(st_polygon(list(rbind(
  c(2000, 2000), c(2800, 2000), c(2800, 2800), c(2000, 2800), c(2000, 2000)
))), crs = 32610)

nrow(create_grid(boundary, cellsize = 150))
nrow(create_grid(boundary, cellsize = 150, exclude = lake))
```

Passing the same layer as `rtm(..., exclude = lake)` removes cells lying *entirely* within the mask from the analysis and the risk map, restricting the estimand to places where events can occur; the remaining zeros are ordinary sampling zeros. Cells only partially covered are kept, since events can still occur in their uncovered part, and a warning is issued if any outcome or denominator events fall inside the mask — a sign of either an over-aggressive mask or bad event coordinates.

Pre-computed cell-level variables (e.g. raster values aggregated to the grid) enter through `covariates`, with an explicit role. A `"candidate"` competes for selection like any risk factor and is reported with an RRV — binary 0/1 values are recommended so the RRV reads as an exposed/unexposed rate ratio. An `"adjustment"` is a confounder held constant: it is included in every candidate model, unpenalized in the culling step, allowed either coefficient sign, and reported separately from the RRV table — continuous values are fine (and preferable) there, and categorical values (factor or character, e.g. a land-use class per cell) are dummy-coded against a `reference` level of your choice, each level's coefficient being the adjusted log rate ratio versus the reference.

```{r covariates, eval = FALSE}
covariates = list(
  greenspace  = list(values = green_bin, role = "candidate", type = "protective"),
  deprivation = list(values = dep_index, role = "adjustment"),
  landuse     = list(values = landuse_class, role = "adjustment", reference = "residential")
)
```

Candidates must be directional, so a categorical covariate cannot be a candidate as-is; hypothesize a specific class with a binary indicator instead (`values = as.integer(landuse_class == "commercial")`).

## How spatial influence is modeled

Spatial influence is **not a fitted decay curve**. It is estimated by *discrete model selection over a ladder of candidate distances*. With `block_length = 300` and `increment = "half"`, each factor is tested at 150, 300, 450, 600, 750 and 900 m, and each candidate variable encodes a different hypothesis about the factor's reach:

-   **Proximity at *d***: a cell is exposed (1) if its centroid lies within *d* of *any* feature — a uniform "step" kernel, so a bar 10 m away and a bar 290 m away count the same under *d* = 300.
-   **Density at bandwidth *d***: a cell is exposed if it sits in a high-*concentration* area — an Epanechnikov kernel density with bandwidth *d* at least 2 standard deviations above the mean. This captures clustering ("bar districts") rather than nearness to a single feature.

The culling and stepwise stages then pick **at most one** variable per factor. Reading "bars / proximity / 300" therefore means: among all tested reaches and operationalizations of bars — and the option of excluding bars entirely — the 300 m proximity coding best explained the cell counts after BIC's complexity penalty.

Three caveats for interpretation:

-   Each factor's influence is a property of the factor *layer*: all bars share one estimated reach. For heterogeneous features (nightclubs vs. pubs), split the layer into separate factors.
-   Risk is flat inside the influence zone and zero outside — a deliberate RTM simplification that keeps the output actionable ("patrol within 300 m of bars").
-   The estimate is only as fine as the ladder: "300 m" means "better than 150 m and 450 m by BIC", and nothing beyond `block_length * max_blocks` is ever considered.

## Interpreting RRVs and risk scores

The final model is a log-linear count regression, so effects are multiplicative:

$$E(\text{count}_j) = e^{\beta_0} \times \prod_i RRV_i^{\,x_{ij}}$$

-   **RRV (relative risk value)** is the exponentiated coefficient — for an epidemiological audience, an *adjusted incidence rate ratio*: cells within the factor's influence zone have RRV times the expected event rate of unexposed cells, holding the other selected factors constant. For protective factors (coded 0/−1), exposed cells have their expected rate *divided* by the RRV (equivalently, an IRR of 1/RRV).
-   **The relative risk score** of a cell is its predicted event count divided by the smallest predicted count in the study area. A score of 10 means "ten times the expected event rate of the safest cell". It is a relative index anchored at 1 — not a probability, and (without an offset; see below) events per unit *area*, not per capita.
-   **p-values are a sanity check, not inference.** Selection is driven by cross-validation and BIC; the significance filter (≤ 0.05) makes the procedure robust, but post-selection p-values should not be reported as ordinary hypothesis tests. For honest post-selection inference see `vignette("post-selection")`.

## Mapping and export

`plot()` maps the risk terrain directly; `rtm_raster()` and `write_rtm()` (requiring the terra package) convert it to a raster or GeoTiff for GIS software, mirroring RTMDx's "output_score" and "output_prediction" files.

```{r map}
plot(fit)
```

```{r export, eval = requireNamespace("terra", quietly = TRUE)}
tif <- file.path(tempdir(), "risk_map.tif")
write_rtm(fit, tif, overwrite = TRUE)
```

## Validating against holdout events

Does the risk terrain predict *future* events? Fit on period 1, then validate on period 2 events the model never saw:

```{r holdout}
set.seed(99)
crimes2 <- do.call(rbind, lapply(seq_along(bars), function(i) {
  b <- st_coordinates(bars[i])
  n <- rpois(1, 8)
  cbind(rnorm(n, b[1], 100), rnorm(n, b[2], 100))
}))
crimes2 <- crimes2[crimes2[, 1] >= 0 & crimes2[, 1] <= size &
  crimes2[, 2] >= 0 & crimes2[, 2] <= size, ]
crimes2 <- as_sfc(crimes2)

val <- validate_rtm(fit, crimes2)
val
```

The capture table reports the share of holdout events falling in the riskiest 1–20% of cells; the Predictive Accuracy Index (PAI) divides that by the share of area flagged, so PAI = 4 means flagged cells catch events at four times the study-area-wide rate. The AUC treats each cell as a binary case (any holdout event vs. none) with the risk score as classifier.

```{r roc}
plot(val)
```

## Rate models: an offset denominator

Passing `offset` turns the count model into a rate model — the natural frame for many public health applications. The example below mimics a fatal-overdose study: overdose events cluster around bars, but an event's probability of being *fatal* only depends on distance to a naloxone distribution site. Using **all overdose events as the offset**, the model estimates *conditional lethality* — case fatality — rather than where overdoses happen.

```{r overdose-data}
set.seed(7)
naloxone <- rand_pts(8)

gen_period <- function() {
  ev <- do.call(rbind, lapply(seq_along(bars), function(i) {
    b <- st_coordinates(bars[i])
    n <- rpois(1, 40)
    cbind(rnorm(n, b[1], 150), rnorm(n, b[2], 150))
  }))
  ev <- rbind(ev, rand_pts(150))
  ev <- ev[ev[, 1] >= 0 & ev[, 1] <= size & ev[, 2] >= 0 & ev[, 2] <= size, ]
  d_nalox <- apply(ev, 1, \(e) min(sqrt((naloxone[, 1] - e[1])^2 + (naloxone[, 2] - e[2])^2)))
  fatal <- runif(nrow(ev)) < ifelse(d_nalox <= 600, 0.03, 0.35)
  list(all = as_sfc(ev), fatal = as_sfc(ev[fatal, , drop = FALSE]))
}

period1 <- gen_period()
period2 <- gen_period() # for holdout validation
```

```{r overdose-fit}
set.seed(3)
fatality <- rtm(
  outcome = period1$fatal,
  offset = period1$all, # fatal + nonfatal events as the denominator
  factors = list(
    bars = bars,
    naloxone = list(data = as_sfc(naloxone), type = "protective")
  ),
  boundary = boundary,
  cell_size = 150,
  block_length = 300,
  operation = "proximity",
  verbose = FALSE
)
fatality
```

The offset absorbs event *volume*, so bars — which drive where overdoses occur but not how deadly they are — are correctly rejected, while the naloxone factor is selected as protective: within its influence zone the case-fatality rate is divided by its RRV. In a rate model, `relrisk` is the relative rate multiplier, defined for *every* cell from its exposures ("if an overdose happened here, how much deadlier than in the safest cell"), while cells with a zero denominator are excluded from fitting — harmlessly, since a place with no events carries no information about lethality.

### Denominator-aware validation

For a rate model, area-based capture is misleading: a high-lethality cell with few overdoses *should* capture few fatal events. Supplying the holdout period's denominator makes the validation denominator-aware:

```{r overdose-validate}
val_fatal <- validate_rtm(fatality, period2$fatal, new_offset = period2$all)
val_fatal
```

`adj_pai` compares the share of fatal events captured with the share of the *denominator* in the same cells: values above 1 mean overdoses occurring in the flagged cells were disproportionately fatal — the model ranks lethality, not just volume. Note how the two metrics can disagree: the plain (area-based) `pai` may fall below 1 because high-lethality cells are not where overdoses are most frequent, while `adj_pai` stays above 1 — which is exactly the property a case-fatality model should have. The observed/expected ratio is computed against the model's expected counts given the holdout denominator, a genuine calibration check.

### Selection-clean coefficients

The RRVs above were *selected* — the pipeline kept naloxone because it looked strong — so their p-values are not ordinary tests, and the estimates themselves are biased away from the null (the winner's curse). RTMR provides two strategies for honest post-selection statements: re-estimating the frozen specification on events that played no part in selection (`split_events()` + `refit_rtm()`, also usable with a later period to check that effects hold prospectively), and bootstrapping the whole selection pipeline when events are too scarce to split (`stability_rtm()`, reporting selection frequencies and selection-aware bagged RRVs on the full sample). Both are covered in `vignette("post-selection")`.

### Sensitivity analysis with a smoothed offset

Raw per-cell denominators are statistically exact, but you may wonder whether sparse denominators (single-event cells) affect the results. `smooth_offset()` builds a kernel-smoothed, total-preserving denominator surface for sensitivity analysis. Smoothing fabricates denominators where nothing was observed and blurs the denominator across exposure boundaries, so it should not be the primary analysis — refit at a couple of bandwidths and check that the RRVs are stable:

```{r overdose-smooth}
grid <- create_grid(boundary, cellsize = 150)

rrv_at <- function(bw) {
  off <- smooth_offset(period1$all, grid, bandwidth = bw)
  set.seed(3)
  refit <- suppressWarnings(rtm(
    outcome = period1$fatal, offset = off,
    factors = list(
      bars = bars,
      naloxone = list(data = as_sfc(naloxone), type = "protective")
    ),
    boundary = boundary, cell_size = 150, block_length = 300,
    operation = "proximity", verbose = FALSE
  ))
  refit$selected[, c("factor", "operation", "spatial_influence", "rrv")]
}

rrv_at(450)
rrv_at(900)
```

The comparison is diagnostic in both directions. Here the naloxone effect is stable across bandwidths (a robustness result worth reporting), but the smoothed models *spuriously select bars* — with an RRV that grows with the bandwidth. That is the numerator–denominator mismatch at work: smoothing moves denominator mass away from the activity peaks around bars while the fatal counts stay put, manufacturing apparent lethality exactly where the denominator was blurred the most. The raw-count model, which rejected bars, is the honest analysis; the smoothed runs bound how sensitive each factor is to denominator noise.

## Notes and caveats

-   **Ecological design.** Units are places, not people; associations are place-level and say nothing about individuals.
-   **Denominators.** Without an offset, risk is per unit area. With an offset it is per denominator unit — and the denominator's own ascertainment (e.g. only *detected* nonfatal overdoses) shapes the estimand.
-   **Association, not causation.** Selected factors are markers of risky environments that survived model selection, not causal effects.
-   **Coverage.** Risk factor layers should include features slightly beyond the study area boundary, so their influence on edge cells is captured.
-   **Reproducibility and fold stability.** The culling step assigns cells to cross-validation folds at random (as in RTMDx, where the assignment changes on every run). Call `set.seed()` before `rtm()` for exact reproducibility. With borderline signals, which variables survive culling can depend on the fold draw; `cull_repeats = 5` (or more) averages the cross-validation over several independent fold assignments, making the culling insensitive to any single draw at proportional extra compute. The default (1) replicates RTMDx.

## References

-   Caplan, J. M., Kennedy, L. W., & Piza, E. L. (2013). *Risk Terrain Modeling Diagnostics Utility User Manual (Version 1.0)*. Newark, NJ: Rutgers Center on Public Security.
-   Heffner, J. (2013). Statistics of the RTMDx Utility. In the above manual.
-   Chainey, S., Tompson, L., & Uhlig, S. (2008). The utility of hotspot mapping for predicting spatial patterns of crime. *Security Journal*, 21, 4–28.
